HTW Model e1

### 2024-03-26 Model/htw_model.qmd
group exam vs alm figure

The residuals of the model predictions for the testing stage (@fig-htw-resid-pred) also show an unsurprising pattern across fitting methods - with models fit only to the test data showing the best performance, followed by models fit to both training and test data, and with models fit only to the training data showing the worst performance (note that y axes are scaled different between plots). Unsurprisingly, the advantage of EXAM is strongest for extrapolation positions (the three smallest bands for both groups - as well as the two highest bands for the Constant group). Although EXAM tends to perform better for both Constant and Varied participants (see also @fig-ee-e1), the relative advantage of EXAM is generally larger for the Constant group - a pattern consistent across all three fitting methods.

@fig-cm-vx-pat directly compares the aggregated observed data to the posterior predictive distributions for the testing stage. Of interest are a) the extent to which the median estimates of the ALM and EXAM posteriors deviate from the observed medians for each velocity band; b) the ability of ALM and EXAM to discriminate between velocity bands; c) the relative performance of models that are constrained by the training data (i.e. the 'fit to train only' and 'fit to both' models) compared to the 'fit to test only' models;

Considering first the models fit to only the testing data, which reflect the best possible performance of ALM and EXAM at capturing the group-aggregated testing patterns. For the varied group, both ALM and EXAM are able to capture the median values of the observed data within the 66% credible intervals, and the spread of model predictions generally matches that of the observed data. For the constant group, only EXAM is able to capture the median range of values across the velocity bands, with ALM generally underestimating human velocoties in the upper bands, and overestimating in the lower bands. In the case of band 100, the median ALM prediction appears to match that of our participants - however this is due to a large subset of participants have ALM predictions near 0 for band 100, a pattern we will explore further in our considertation of individual patterns below. Models fit to both training and testing data show a similar pattern to only the testing data display the same basic pattern as those fit to only the testing data, albeit with slightly larger residuals. However models fit to only the training data display markedly worse performance at accounting for the key testing patterns.

-   \*\* explain how the constant group ALM predictions for band 100 look deceptively good due to aggregation of a large subset of subjects having ALM predictions of 0 for vb100, and a large subset with ALM predictions close to their position 800 value. This is relected by much greater variance of the ALM esimates in the posterior predictive plot

-   \*\* comment on how much constrained by the training data has a worse impact on the EXAM predictions for varied than for constant - perhaps due to the varied training data being much noisier than the constant training data.

-   \*\* comment on EXAM doing a better job mimicing the within-condition variance of the observed data

-   \*\* comment on the % of Constant subjects being best accounted for by EXAM being higher.

@@

@@

# {btw_con |> ggplot(aes(x=value,y=Model,fill=Model)) +stat_halfeye()} +
# {btw_model |> ggplot(aes(x=value,y=contrast,fill=condit)) +stat_halfeye()}

# emm_models_bandint_specific <- e1_ee_brm_ae |>
#   emmeans(~ Model | condit, at = list(bandInt = c(100,350,600,800,1000,1200)),
#           re_formula = NA)
# 
# 
# emm_models_bandint_by_levels <- e1_ee_brm_ae |>
#   emmeans(~ Model | condit | bandInt, 
#           at = list(bandInt = c(100,350,600,800,1000,1200)),
#           re_formula = NULL, by = "bandInt")

# within_condition_bandint_by_levels_contrasts <- emm_models_bandint_by_levels |>
#   pairs()
# 
# emm_models_bandint_by_levels |> as_tibble()

# emm_models_bandint_by_levels |> pairs() |> gather_emmeans_draws() |> 
#   as_tibble() |> group_by(bandInt,condit) |> tidybayes::median_hdi(.value) |>
#   ggplot(aes(x=.value,y=condit,fill=condit)) + stat_halfeye() +facet_wrap(~bandInt,ncol=1)

#print(summary_EXAM)

# alternative emmeans approach to assess constant vs. varied difference in EXAM
#emmeans(e1_ee_brm_ae, specs = "condit", by = "Model") |> summary()
# is the difference significant for EXAM in particular

# and errors for the Varied condition than the Constant condition ($\beta$ = `r bm2$estimate`, 95% CrI \[`r bm2$conf.low`, `r bm2$conf.high`\], pd = `r bm2$pd`).

@@

To quantitatively assess whether the differences in performance betwen ALM,  we fit a bayesian regressions predicting the errors of the posterior predictions of each models as a function of the Model (ALM vs. EXAM) and training condition (Constant vs. Varied).

@@

```{r}
#| label: fig-cm-vx-pat
#| fig-cap: "Empirical data and Model predictions for mean velocity. Observed data is shown on top, and predictions from each combination of model and fitting method are shown below. Bolded bars indicate bands that were trained, non-bold bars indicate extrapolation bands. "
#| fig-height: 10
#| fig-width: 10

#post_dat_l |> group_by(id,condit,x) |> filter(Resp=="Observed") |> slice_head(n=1)

pemp1 <- e1 |> filter(expMode2=="Test") |> mutate(Resp="Observed") |> 
  ggplot(aes(x=condit,y=vx, fill=vb, col=ifelse(bandType=="Trained","black",NA),size=ifelse(bandType=="Trained","black",NA))) + 
  stat_bar+ 
  scale_color_manual(values = c("black" = "black"), guide = FALSE) +
  scale_size_manual(values = c("black" = .5), guide = FALSE) + 
  theme(legend.position="right", 
  axis.title.x = element_blank(), 
  plot.margin = unit(c(0,0,0,0), "pt"),
  plot.title = element_text(hjust=.50)) +
  labs(title="Empirical Data - Experiment 1",y="vx", x="Condition",fill="Band")

layout <- "
#A#
CCC
"

pmod1 <- post_dat_l |> filter(!(Resp=="Observed")) |> 
  group_by(id,condit, Fit_Method,Resp,bandType,x,vb) |> 
 summarize(vx=median(val)) |> 
 #left_join(testAvgE1, by=join_by(id,condit,x==bandInt)) |>
 ggplot( aes(x=condit,y=vx, fill=vb,col=ifelse(bandType=="Trained","black",NA),size=ifelse(bandType=="Trained","black",NA))) + 
  stat_bar + 
    facet_wrap(~Resp+rename_fm(Fit_Method), strip.position = "top", scales = "free_x") +
        scale_color_manual(values = c("black" = "black"), guide = "none") +
  scale_size_manual(values = c("black" = .5), guide = "none") +
    theme(panel.spacing = unit(0, "lines"), 
         strip.background = element_blank(),
         strip.placement = "outside",
         legend.position = "none",plot.title = element_text(hjust=.50),
         axis.title.x = element_blank(),
         plot.margin = unit(c(10,0,0,0), "pt")) + 
         labs(title="Model Predictions - Experiment 1 Data", y="vx")

# (pemp1  / pmod1 ) + 
#   plot_layout(design = layout, heights = unit(c(6,6),c('cm','cm'))) #heights = unit(c(5,-5, 8), c('cm','null'))

pemp1/pmod1 + 
  plot_layout(design = layout, 
  heights = c(3,6),
  widths = c(5,5))

```{r}
#| label: fig-cm-dev-pat
#| fig-cap: "Empirical data and Model predictions for absolute deviation from target (accuracy). Observed data is shown on top, and predictions from each combination of model and fitting method are shown below. "
#| fig-height: 11
#| fig-width: 11
#| eval: false

pemp1 <- e1 |> filter(expMode2=="Test") |> 
  group_by(id,condit,vb,bandType) |> 
  ggplot(aes(x=condit,y=dist, fill=vb, col=ifelse(bandType=="Trained","black",NA),size=ifelse(bandType=="Trained","black",NA))) + 
  stat_summary(fun=mean, geom="bar", position=position_dodge()) + 
  stat_summary(fun.data=mean_se, geom="errorbar", color="black", position=position_dodge(), size=.5) + 
  scale_color_manual(values = c("black" = "black"), guide = FALSE) +
  scale_size_manual(values = c("black" = .5), guide = FALSE) + 
  scale_y_continuous( breaks=seq(0,600,by=200),labels=as.character(seq(0,600,by=200))) +
  expand_limits(y=600) +
  theme(legend.position="right", axis.title.x = element_blank(),plot.title = element_text(hjust=.40)) +
  labs(title="Empirical Data", y="Deviation from target band", fill="Band")

# layout <- "#A#
# #B#
# CCC"

layout <- "#A#
CCC"

pmod1 <- post_dat_l |> filter(!(Resp=="Observed")) |> 
  #left_join(testAvgE1, by=join_by(id,condit,x==bandInt)) |>
  group_by(id,condit, Fit_Method,Resp,vb,bandType) |> 
 summarize(dist=median(dist)) |> 
 ggplot( aes(x=condit,y=dist, fill=vb,col=ifelse(bandType=="Trained","black",NA),size=ifelse(bandType=="Trained","black",NA))) + 
  stat_bar + 
    facet_wrap(~Resp+rename_fm(Fit_Method), strip.position = "top", scales = "free_x") +
        scale_color_manual(values = c("black" = "black"), guide = FALSE) +
  scale_size_manual(values = c("black" = .5), guide = FALSE) +
    theme(panel.spacing = unit(0, "lines"), 
         strip.background = element_blank(),
         strip.placement = "outside",
         legend.position = "none",
         plot.title = element_text(hjust=.50),
         plot.margin = unit(c(20,0,0,0), "pt"),
         axis.title.x = element_blank()) + 
         labs(title="Model Predictions - Experiment 1 Data", y="Deviation from target band")

(pemp1 / pmod1) + plot_layout(design = layout) #heights = unit(c(5,-5, 8), c('cm','null'))

```{r}
#| label: tbl-htw-modelError-e123
#| tbl-cap: "Results of Bayesian Regression models predictin model error as a function of Model (ALM vs. EXAM), Condition (Constant vs. Varied), and the interaction between Model and Condition. The values represent the estimate coefficient for each term, with 95% credible intervals in brackets.  The intercept reflects the baseline of ALM and Constant. The other estimates indicate deviations from the baseline for the EXAM mode and varied condition. Lower values indicate better model fit."

@@

e2_tab <- rbind(post_tabs2$agg_pred_full |> mutate("Task Stage"="Test"), train_tab2$agg_pred_full |> mutate("Task Stage"="Train")) |> 
  mutate(Fit_Method=rename_fm(Fit_Method)) 
  # |>
  # flextable::tabulator(rows=c("Task Stage","Fit_Method","Model"), columns=c("condit"),
  #                      `ME` = as_paragraph(mean_error)) |> as_flextable()

bm1 <- get_coef_details(e2_ee_brm_ae, "conditVaried")
bm2 <- get_coef_details(e2_ee_brm_ae, "ModelEXAM")
bm3 <- get_coef_details(e2_ee_brm_ae, "ModelEXAM:conditVaried")
# bct_e2 <- as.data.frame(bayestestR::describe_posterior(e2_ee_brm_ae, centrality = "Mean")) %>%
#   select(1,2,4,5,6) %>%
#   setNames(c("Term", "Estimate","95% CrI Lower", "95% CrI Upper", "pd")) %>%
#   mutate(across(where(is.numeric), \(x) round(x, 2))) %>%
#   tibble::remove_rownames() %>%
#   mutate(Term = stringr::str_remove(Term, "b_")) %>%
#   kable(booktabs = TRUE)

e3_model <- load_e3()
post_tabs3 <- abc_tables(e3_model$post_dat,e3_model$post_dat_l)
train_tab3 <- abc_train_tables(e3_model$pd_train,e3_model$pd_train_l)

@@

# e3_ee_brm_ae <- brm(data=pdl3,
#   aerror ~  Model * condit+ (1+bandInt|id) + (1+condit|bandOrder), 
#   file = paste0(here("data/model_cache/e3_ae_modelCondBo_RFint.rds")),
#   chains=4,silent=1, iter=2000, control=list(adapt_delta=0.92, max_treedepth=11))

@@

bm1 <- get_coef_details(e3_ee_brm_ae, "conditVaried")
bm2 <- get_coef_details(e3_ee_brm_ae, "ModelEXAM")
bm3 <- get_coef_details(e3_ee_brm_ae, "ModelEXAM:conditVaried")

@@

>>>>

The residuals of the model predictions for the testing stage (@fig-htw-resid-pred) also show an unsurprising pattern across fitting methods - with models fit only to the test data showing the best performance, followed by models fit to both training and test data, and with models fit only to the training data showing the worst performance (note that y axes are scaled different between plots). Although EXAM tends to perform better for both Constant and Varied participants (see also @fig-ee-e1), the relative advantage of EXAM is generally larger for the Constant group - a pattern consistent across all three fitting methods. The primary predictive difference between ALM and EXAM is made clear in @fig-cm-vx-pat, which directly compares the observed data against the posterior predictive distributions for both models. Regardless of how the models are fit, only EXAM can capture the pattern where participants are able to discriminate all 6 target bands.

@@

```{r}
#| label: fig-cm-vx-pat
#| fig-cap: "Empirical data and Model predictions for mean velocity across target bands. Fitting methods (Test Only, Test & Train, Train Only) - are separated across rows, and Training Condition (Constant vs. Varied) are separated by columns. Each facet contains the predictions of ALM and EXAM, alongside the observed data."
#| fig-height: 10
#| fig-width: 10

post_dat_l |> 
  group_by(id,condit, Fit_Method,Resp,bandType,x,vb) |> 
 summarize(vx=median(val)) |> 
 #left_join(testAvgE1, by=join_by(id,condit,x==bandInt)) |>
 ggplot( aes(x=Resp,y=vx, fill=vb,col=ifelse(bandType=="Trained","black",NA),size=ifelse(bandType=="Trained","black",NA))) + 
  stat_bar + 
    facet_wrap(~rename_fm(Fit_Method)+condit, ncol=2,strip.position = "top", scales = "free_x") +
        scale_color_manual(values = c("black" = "black"), guide = "none") +
  scale_size_manual(values = c("black" = .5), guide = "none") +
    theme(panel.spacing = unit(0, "lines"), 
         strip.background = element_blank(),
         strip.placement = "outside",
         legend.position = "none",plot.title = element_text(hjust=.50),
         axis.title.x = element_blank(),
         plot.margin = unit(c(10,0,0,0), "pt")) + 
         labs(title="Model Predictions - Experiment 1 Data", y="Vx")
```

@@

@@

To quantitatively assess whether the differences in performance between models,  we fit a bayesian regressions predicting the errors of the posterior predictions of each models as a function of the Model (ALM vs. EXAM) and training condition (Constant vs. Varied).

@@

@@

e2_tab <- rbind(post_tabs2$agg_pred_full |>
 mutate("Task Stage"="Test"), train_tab2$agg_pred_full |> 
 mutate("Task Stage"="Train")) |> 
  mutate(Fit_Method=rename_fm(Fit_Method)) 
  # |>
  # flextable::tabulator(rows=c("Task Stage","Fit_Method","Model"), columns=c("condit"),
  #                      `ME` = as_paragraph(mean_error)) |> as_flextable()
e3_model <- load_e3()
post_tabs3 <- abc_tables(e3_model$post_dat,e3_model$post_dat_l)
train_tab3 <- abc_train_tables(e3_model$pd_train,e3_model$pd_train_l)

@@

```{r}
#| label: tbl-htw-ee-e23
#| tbl-cap: "Results of Bayesian Regression models predicting model error as a function of Model (ALM vs. EXAM), Condition (Constant vs. Varied), and the interaction between Model and Condition. The values represent the estimate coefficient for each term, with 95% credible intervals in brackets.  The intercept reflects the baseline of ALM and Constant. The other estimates indicate deviations from the baseline for the EXAM mode and varied condition. Lower values indicate better model fit."

bm1_e2 <- get_coef_details(e2_ee_brm_ae, "conditVaried")
bm2_e2 <- get_coef_details(e2_ee_brm_ae, "ModelEXAM")
bm3_e2 <- get_coef_details(e2_ee_brm_ae, "ModelEXAM:conditVaried")
# bct_e2 <- as.data.frame(bayestestR::describe_posterior(e2_ee_brm_ae, centrality = "Mean")) %>%
#   select(1,2,4,5,6) %>%
#   setNames(c("Term", "Estimate","95% CrI Lower", "95% CrI Upper", "pd")) %>%
#   mutate(across(where(is.numeric), \(x) round(x, 2))) %>%
#   tibble::remove_rownames() %>%
#   mutate(Term = stringr::str_remove(Term, "b_")) %>%
#   kable(booktabs = TRUE)

@@

bm1_e3 <- get_coef_details(e3_ee_brm_ae, "conditVaried")
bm2_e3  <- get_coef_details(e3_ee_brm_ae, "ModelEXAM")
bm3_e3  <- get_coef_details(e3_ee_brm_ae, "ModelEXAM:conditVaried")

@@

```{r}
#| label: fig-htw-best-model
#| fig-cap: Difference in model errors for each participant, with models fit to both train and test data. Positive values favor EXAM, while negative values favor ALM. 
#| fig-height: 12
#| fig-width: 11

tid1 <- post_dat  |> mutate(Exp="E1",bandOrder="Original") |> select(-pred_dist, -dist) |>
  rbind(e2_model$post_dat |> mutate(Exp="E2",bandOrder="Reverse")) |>
  rbind(e3_model$post_dat |> mutate(Exp="E3")) |>
  filter(Fit_Method=="Test_Train") |>
  group_by(id,condit,Model,Fit_Method,x, Exp) |> 
    mutate(e2=abs(y-pred)) |> 
    summarise(y1=median(y), pred1=median(pred),mean_error=abs(y1-pred1)) |>
    group_by(id,condit,Model,Fit_Method,Exp) |> 
    summarise(mean_error=mean(mean_error)) |> 
    arrange(id,condit,Fit_Method) |>
    round_tibble(1)

best_id <- tid1 |> 
  group_by(id,condit,Fit_Method) |> 
  mutate(best=ifelse(mean_error==min(mean_error),1,0))

lowest_error_model <- best_id %>%
  group_by(id, condit,Fit_Method, Exp) %>%
  summarise(Best_Model = Model[which.min(mean_error)],
            Lowest_error = min(mean_error),
            differential = min(mean_error) - max(mean_error)) %>%
  ungroup()

error_difference<- best_id %>%
  select(id, condit, Model,Fit_Method, mean_error) %>%
  pivot_wider(names_from = Model, values_from = c(mean_error)) %>%
  mutate(Error_difference = (ALM - EXAM))

full_comparison <- lowest_error_model |> 
  left_join(error_difference, by=c("id","condit","Fit_Method"))  |> 
  group_by(condit,Fit_Method,Best_Model) |> 
  mutate(nGrp=n(), model_rank = nGrp - rank(Error_difference) ) |> 
  arrange(Fit_Method,-Error_difference)

full_comparison |> 
  filter(Fit_Method=="Test_Train") |> 
  ungroup() |>
  mutate(id = reorder(id, Error_difference)) %>%
  ggplot(aes(y=id,x=Error_difference,fill=Best_Model))+
  geom_col() +
  #ggh4x::facet_grid2(~condit,axes="all",scales="free_y", independent = "y")+
  ggh4x::facet_nested_wrap(~condit+Exp,scales="free") + 
  labs(fill="Best Model",
  x="Mean Model Error Difference (ALM - EXAM)",
  y="Participant")

### 2024-03-26 Model/htw_model_e1.qmd
group exam vs alm figure

>>>>

## Model predictions of deviation

```{r}
#| label: fig-cm-dev-pat
#| fig-cap: "Empirical data and Model predictions for absolute deviation from target (accuracy). Observed data is shown on top, and predictions from each combination of model and fitting method are shown below. "
#| fig-height: 11
#| fig-width: 11
#| eval: false

pemp1 <- e1 |> filter(expMode2=="Test") |> 
  group_by(id,condit,vb,bandType) |> 
  ggplot(aes(x=condit,y=dist, fill=vb, col=ifelse(bandType=="Trained","black",NA),size=ifelse(bandType=="Trained","black",NA))) + 
  stat_summary(fun=mean, geom="bar", position=position_dodge()) + 
  stat_summary(fun.data=mean_se, geom="errorbar", color="black", position=position_dodge(), size=.5) + 
  scale_color_manual(values = c("black" = "black"), guide = FALSE) +
  scale_size_manual(values = c("black" = .5), guide = FALSE) + 
  scale_y_continuous( breaks=seq(0,600,by=200),labels=as.character(seq(0,600,by=200))) +
  expand_limits(y=600) +
  theme(legend.position="right", axis.title.x = element_blank(),plot.title = element_text(hjust=.40)) +
  labs(title="Empirical Data", y="Deviation from target band", fill="Band")

# layout <- "#A#
# #B#
# CCC"

layout <- "#A#
CCC"

pmod1 <- post_dat_l |> filter(!(Resp=="Observed")) |> 
  #left_join(testAvgE1, by=join_by(id,condit,x==bandInt)) |>
  group_by(id,condit, Fit_Method,Resp,vb,bandType) |> 
 summarize(dist=median(dist)) |> 
 ggplot( aes(x=condit,y=dist, fill=vb,col=ifelse(bandType=="Trained","black",NA),size=ifelse(bandType=="Trained","black",NA))) + 
  stat_bar + 
    facet_wrap(~Resp+rename_fm(Fit_Method), strip.position = "top", scales = "free_x") +
        scale_color_manual(values = c("black" = "black"), guide = FALSE) +
  scale_size_manual(values = c("black" = .5), guide = FALSE) +
    theme(panel.spacing = unit(0, "lines"), 
         strip.background = element_blank(),
         strip.placement = "outside",
         legend.position = "none",
         plot.title = element_text(hjust=.50),
         plot.margin = unit(c(20,0,0,0), "pt"),
         axis.title.x = element_blank()) + 
         labs(title="Model Predictions - Experiment 1 Data", y="Deviation from target band")

(pemp1 / pmod1) + plot_layout(design = layout) #heights = unit(c(5,-5, 8), c('cm','null'))
